---
title: "Assessment_6"
format: html
editor: visual
---

## Set up
```{r}
library(readr)
library(heplots)
library(tidyr)
library(dplyr)

# Question 1
skulls <- read.csv("Multivariate_Analysis/Lecture_6/CA1.csv")
head(skulls)

# Question 2
fabric <- read_table("CA6.txt")
head(fabric)
```

## Question 1
```{r}
# Box's M-test 
# Testing if covariance matrices for time periods are equal 
y <- as.matrix(skulls[,1:4], nrow = nrow(skulls))
group <- as.matrix(skulls[, "TimePeriod"])

boxM(y, group) 

```

```{r}
# One-way MANOVA
# Testing whether mean vectors of all time periods are equal 
X <- cbind(skulls$MaxBreadth, skulls$BasHeight, skulls$BasLength, skulls$NasHeight)
group <- as.factor(skulls$TimePeriod)
fit <- manova(X ~ group)
summary(fit, 'Wilks')
```

The p-value from summary is very small. provides evidence that at least one mean vector differs. Reject H0 (that all means are equal across time periods).

## Question 2
```{r}
# Conduct a MANOVA using Pillai's trace
attach(fabric)
X <- cbind(y1, y2, y3)

# Fit two-way MANOVA
fit1 <- manova(X ~ Proportion*Filler)
summary1 <- summary(fit1, test = "Pillai")

# Fit individual ANOVA's 
summary(aov(X ~ Proportion))
summary(aov(X ~ Filler))
```
Question a
2-way MANOVA
Proportion: small p-val = 0.0068
Filler: v small p-val = 4.8e-07
Interaction: small p-val = 0.0049
*look in example 6.13 from notes* 

Individual ANOVA
Proportion:
y1 - 0.0058 
y2 - 0.3604
y3 - 0.545
Becomes less significant over time 

Filler:
y1 - 0.606
y2 - 3.98e-05
y3 - 1.53e-05
Becomes more significant in second and third rev 
not significant in first 

```{r}
# Interaction
V1 <- Proportion
V2 <- Filler

# Plot interactions
par(mfrow=c(3,2))
interaction.plot(V1, V2, X[,1], 
                 ylab = bquote(bar(x)[1]), xlab = 'Factor 1', 
                 col = c('purple', 'deeppink'), trace.label = 'Factor2', leg.bty = 'o')
boxplot(X[,1] ~ V1*V2, col = "steelblue", ylab = expression(x[1]), xlab = 'Factor 1:Factor2')

interaction.plot(V1, V2, X[,2], 
                 ylab = bquote(bar(x)[2]), xlab = 'Factor 1', 
                 col = c('purple', 'deeppink'), trace.label = 'Factor 2', leg.bty = 'o')
boxplot(X[,2] ~ V1*V2, col = "steelblue", ylab = expression(x[2]), xlab = 'Factor1:Factor2')

interaction.plot(V1, V2, X[,3], 
                 ylab = bquote(bar(x)[3]), xlab = 'Factor 1', 
                 col = c('purple', 'deeppink'), trace.label = 'Factor 2', leg.bty = 'o')
boxplot(X[,3] ~ V1*V2, col = "steelblue", ylab = expression(x[3]), xlab = 'Factor1:Factor2')

```

I'm not the biggest fan of how these are set up - idk it looks kinda ugly. I'll try fiddle with it another time and see if I can make it look a bit nicer :) 

Question b
Interaction term: Do the effects of dependent variable 1 on the independent variable depend on the level of dependent variable 2? 

Question c
Proportion is not significant but type of filler is? Not 100% sure how to explain this properly. 
