---
title: "Multivariate Analysis Mock Exam"
subtitle: "STA2005S — Practice Paper (based on breakdown discussion)"
format: html
editor: 
  markdown: 
    wrap: 72
---

# Introduction

```{r}
library(ca)
library(MASS)
library(pls)
library(dplyr)
library(ggplot2)
library(knitr)
```

## Instructions

To start:

1.  Place this file in a project folder alongside a `data/` subfolder
    containing: `pollution_data.csv`, `complaints_data.csv`,
    `yield_data.csv`, `student_data.csv`, `pcr_pls_data.csv`,
    `lda_data.csv`.
2.  Open this `.qmd` in RStudio and render periodically to check it
    works.

## Desired answers

Each question has leading indicators telling you what should be
reflected in your write-up:

-   **R code**: your R code should be typeset (`echo = TRUE`). Where a
    question says "from first principles", you must obtain the output
    using the underlying formulae rather than a wrapper function (e.g.
    not `lm()`, not `lda()`), unless a part explicitly allows a package
    function for cross-checking.
-   **Plot**: display a plot with labelled axes.
-   **Table**: display a table (`kable()` is fine, but not required).
-   **Typed response**: a short paragraph-style answer, typically one
    mark per point made.

## Exam composition

| Section             | Topic                                             | Marks  |
|---------------------|---------------------------------------------------|--------|
| Q1                  | Factor Analysis                                   | 13     |
| Q2                  | Correspondence Analysis                           | 6      |
| Q3.1                | Regression                                        | 5      |
| Q3.2                | Canonical Correlation Analysis (first principles) | 7      |
| Q3.3                | PCA / PCR / PLS                                   | 15     |
| Q3.4                | Fisher's LDA / CDA                                | 12     |
| MCQ                 | Multiple choice                                   | 5      |
| **Total available** |                                                   | **63** |
| **Marked out of**   |                                                   | **60** |

You have five hours to complete the exam.

```{r setup}
library(dplyr)
library(knitr)
library(ggplot2)
```

------------------------------------------------------------------------

# Question 1 --- Factor Analysis (13)

We use `data/pollution_data.csv`: 150 days of measurements on 8
air-quality and weather variables (`no2`, `so2`, `co`, `pm10`, `temp`,
`humidity`, `windspeed`, `pressure`).

```{r}
pollution <- read.csv("data/pollution_data.csv")
head(pollution)
```

a.  **R code + typed response:** Motivate why these data should be
    standardised before factor analysis. Standardise the data. (2)

    This data needs to be standardised as the data is on a lot of
    different scales froe example temp is in celsius while things like
    no2 are in a completely different type of measurement. Pressure is
    also an incredibly high values compared to the other variables and
    we dont want that to play a part in our model, so we standardise.

    ```{r}
    col_means = colMeans(pollution)
    col_sd = apply(pollution, 2, sd)
    pol_scaled = scale(pollution)
    ```

b.  **R code + table:** From first principles, using the correlation
    matrix and its eigendecomposition, extract the **first factor only**
    via the principal-component method. Report its loadings, the
    communalities, and the proportion of total variance it explains. (4)

    ```{r}
    pol_cor_mat = cor(pol_scaled)
    eig_fa = eigen(pol_cor_mat)
    prop_pc = sum(eig_fa$values[1]) / sum(eig_fa$values)
    prop_pc

    L_pc = eig_fa$vectors %*% diag(sqrt(eig_fa$values))
    LL_pc = L_pc[,1] %*% t(L_pc[,1])
    kable(data.frame(Loading1 = LL_pc[,1]))

    h = diag(LL_pc)
    h

    ### Extra for completeness
    psi = 1 - h
    cor_recreate = LL_pc + diag(psi)
    kable(cor_recreate)
    ```

c.  **R code:** Fit a two-factor model to the standardised data using
    the ML approach (`factanal`). (2)

    ```{r}
    FA = factanal(covmat = pol_cor_mat, factors = 2, rotation = "none")
    L_ml = FA$loadings
    L_ml

    ## For extra!!

    # FA_rot = varimax(FA$loadings)
    # FA_rot
    # 
    # PC_rot = varimax(L_pc[,1:2])
    # PC_rot
    ```

d.  **R code + table:** From first principles, calculate Bartlett's WLS
    factor scores for the two-factor ML solution above, i.e.
    $$\hat{\mathbf{F}} = (\hat{\mathbf{L}}'\hat{\boldsymbol{\Psi}}^{-1}\hat{\mathbf{L}})^{-1}\hat{\mathbf{L}}'\hat{\boldsymbol{\Psi}}^{-1}(\mathbf{x}-\bar{\mathbf{x}})$$
    Display the first six rows of scores. (4)

    ```{r}
    psi = FA$uniquenesses
    psi_inv = diag(1/psi)
    A = solve(t(L_ml) %*% psi_inv %*% L_ml)
    B = A %*% t(L_ml) %*% psi_inv
    Z = scale(pollution)

    F_scores = Z %*% t(B)
    kable(head(F_scores), digits = 4)

    ## for extra, reconstruct data from afctor scores
    reconst = (F_scores %*% t(L_ml)) * col_sd + col_means
    kable(head(reconst))
    kable(head(reconst == pollution), digits = 4)
    ```

e.  **Typed response:** Based on the loadings, suggest a name for each
    factor. (1)

    no

------------------------------------------------------------------------

# Question 2 --- Correspondence Analysis (6)

We use `data/complaints_data.csv`: complaint counts (Billing, Staff,
Product, Delivery) logged at four store branches.

```{r}
complaints <- read.csv("data/complaints_data.csv", row.names = 1)
kable(complaints)
```

a.  **R code + plot:** Produce a symmetric CA biplot of this table. (2)

    ```{r}
    ca_obj = ca(complaints)
    plot(ca_obj, map = "symmetric")

    plot(ca_obj, map = "rowprincipal", contrib = "absolute", arrows = c(TRUE, TRUE), mass = TRUE)
    ```

b.  **Typed response:** Interpret the plot. Use language such as "drives
    separation along" or "contributes to inertia" rather than
    "associated with". (2)

    Since this is a symmetric biplot and both cols and rows are plotted
    using principal coordinates we can only compare rows to rows and
    cols to cols no inter comparison. When looking at the rows dim 1
    which explains 47% of the inertia drives seperation for observatory
    and rondebosch from claremont and bellville. Dim2 which explains 35%
    of the inertia seperates Claremont and Rondebosch from Observatory
    and Bellville. Similar stuff can be said about the cols. Obs and
    Product are closets to the origin and thus have the least deviation
    from what they would look like under independance.

c.  **R code + table:** Using `summary(ca(complaints))`, identify which
    branch contributes most to the inertia of the first dimension, and
    report its mass, quality (`qlt`), and contribution (`ctr`) on that
    dimension. (2)

    its claremont with a ctr of 463.

    ```{r}
    summary(ca_obj)
    ```

------------------------------------------------------------------------

# Question 3

## 3.1 --- Regression (5)

We use `data/yield_data.csv`: crop yield with soil pH and rainfall as
predictors.

```{r}
yield_dat <- read.csv("data/yield_data.csv")
head(yield_dat)
```

a.  **R code + table:** From first principles, fit a multivariate...
    (here, simple) linear regression of `yield` on `soil_ph` and
    `rainfall`. Display the estimated regression coefficients. (2)

    ```{r}
    Y = cbind(yield_dat$yield)
    X = cbind(1, yield_dat$soil_ph, yield_dat$rainfall)
    X_inv = solve(t(X) %*% X)
    Bhat = X_inv %*% t(X) %*% Y
    Yhat = X %*% Bhat

    kable(data.frame(vars = c("intercept", "soil", "rainfall"),estimates = Bhat), digits = 4)
    ```

b.  **R code + output:** From first principles, construct a 95%
    confidence interval for the slope on `soil_ph`. (2)

    ```{r}
    n = nrow(Yhat)
    k = ncol(X)
    s = 1 / (n - k) * t(Y - Yhat) %*% (Y - Yhat)
    se = sqrt(s * X_inv[2,2])
    t_crit = qt(0.975, df = n - k)
    Upper = Bhat[2,1] + t_crit * se
    Lower = Bhat[2,1] - t_crit * se
    kable(data.frame(Upper = Upper, Lower = Lower, estimate = Bhat[2,1]))
    ```

c.  **Typed response:** Interpret the interval. (1)

    no

## 3.2 --- Canonical Correlation Analysis (7)

We use `data/student_data.csv`: an "academic" set of variables
($X^{(1)}$: `gpa`, `test_score`, `attendance`, `study_hrs`) and a
"lifestyle" set ($X^{(2)}$: `sleep_hrs`, `exercise_hrs`, `stress`).

```{r}
student <- read.csv("data/student_data.csv")
head(student)
```

a.  **R code + output:** From first principles, obtain the first
    canonical correlation and the associated coefficient vectors
    $\mathbf{a}_1$, $\mathbf{b}_1$. (3)

    ```{r}
    cor_mat = cor(student)
    #cor_mat
    Cor11 = cor_mat[1:4,1:4]
    Cor22 = cor_mat[5:7,5:7]
    Cor12 = cor_mat[1:4,5:7]
    Cor21 = cor_mat[5:7,1:4]

    rho11_inv = solve(Cor11)
    eig = eigen(rho11_inv)
    rho11_inv_sqrt = eig$vectors %*% diag(sqrt(eig$values)) %*% t(eig$vectors)
    rho22_inv = solve(Cor22)

    R = rho11_inv_sqrt %*% Cor12 %*% rho22_inv %*% Cor21 %*% rho11_inv_sqrt
    eigR = eigen(R)
    canonical_cor = sqrt(eigR$values)[1]
    canonical_cor

    a1 = (rho11_inv_sqrt %*% eigR$vectors)[,1]

    b_1 = rho22_inv %*% Cor21 %*% a1
    standardiser = sqrt(diag(t(b_1) %*% Cor22 %*% b_1))
    b1 = b_1 / standardiser

    kable(a1)
    kable(b1)

    # Check
    X1 = student[,1:4]
    X2 = student[,5:7]
    test = cancor(X1,X2)
    test

    ```

b.  **R code + output:** Compute the canonical loadings and
    cross-loadings for the first pair of canonical variates. (2)

    ```{r}
    # DONT FORGET TO SCALE!!!
    X1_scaled = scale(X1)
    X2_scaled = scale(X2)
    U1 = as.matrix(X1_scaled) %*% a1
    V1 = as.matrix(X2_scaled) %*% b1

    U1X1 = cor(U1,X1_scaled)
    V1X2 = cor(V1,X2_scaled)
    kable(data.frame(U1X1))
    kable(data.frame(V1X2))

    U1X2 = cor(U1,X2_scaled)
    V1X1 = cor(V1,X1_scaled)
    kable(data.frame(U1X2))
    kable(data.frame(V1X1))
    ```

c.  **R code + output:** Verify your first canonical correlation using
    `cancor()`. (2)

    Done above

## 3.3 --- PCA / PCR / PLS (15)

We use `data/pcr_pls_data.csv`: a response `y` and ten correlated
predictors `x1`--`x10`.

```{r}
pcr_dat <- read.csv("data/pcr_pls_data.csv")
head(pcr_dat)
```

a.  **R code + plot:** Standardise the predictors, perform PCA from
    first principles, and produce a scree plot. State how many
    components you would retain. (3)

    ```{r}
    dat_scaled = scale(pcr_dat[,1:10])
    pc_cor = cor(dat_scaled)
    eig_pc = eigen(pc_cor)
    prop = eig_pc$values / sum(eig_pc$values)
    plot(prop, type = "o")
    # Select 4 PCs
    loadings = eig_pc$vectors[,1:4]
    kable(loadings)
    scores = dat_scaled %*% loadings
    kable(scores)

    # svd = svd(dat_scaled)
    # prop_alt = svd$d^2 / sum(svd$d^2)
    # plot(prop_alt, type = "o")
    # loadings_alt = svd$v[,1:4]
    # kable(loadings_alt)
    # scores_1 = dat_scaled %*% loadings_alt
    # kable(scores_1)
    # scores_2 = (svd$u %*% diag(svd$d))[,1:4]
    # kable(scores_2)
    ```

b.  SSE

c.  **R code + table:** From first principles, fit principal component
    regression (PCR) of `y` on your retained components. Report $R^2$.

    (4) 

    ```{r}
    X = cbind(1, scores)
    Y = as.matrix(scale(pcr_dat[,11]))
    X_inv = solve(t(X) %*% X)
    Bhat_pc = X_inv %*% t(X) %*% Y
    Yhat = X %*% Bhat_pc

    SSE = t(Y) %*% Y - t(Yhat) %*% Yhat
    n = nrow(Yhat)
    ybar = mean(Y)
    SST = t(Y) %*% Y - n * ybar^2
    R_pcr = 1 - SSE / SST
    R_pcr

    ```

d.  **R code + table:** Fit a partial least squares (PLS) regression of
    `y` on the ten predictors using the `pls` package, with the same
    number of components. Report $R^2$. (4)

    ```{r}
    library(pls)
    Y = as.matrix(pcr_dat[,11])
    df = data.frame(pcr_dat)
    pls_fit = plsr(y ~ ., data = df, validation = "CV", scale = TRUE)
    validationplot(pls_fit, val.type = "MSEP")
    bncomp = 4

    Bhat_pls = coef(pls_fit, ncomp = bncomp, intercept = TRUE)
    yhat_pls = as.matrix(predict(pls_fit, ncomp = bncomp, newdata = df))

    SSE = t(Y) %*% Y - t(yhat_pls) %*% yhat_pls
    ybar = mean(Y)
    n = nrow(Yhat)
    SST = t(Y) %*% Y - n * ybar^2

    R_pls = 1 - SSE / SST
    R_pls
    ```

e.  **Typed response:** Compare the $R^2$ values from PCR, PLS, and a
    full multiple regression using all ten predictors. Explain why PLS
    typically needs fewer components than PCR to achieve comparable
    predictive performance. (4)

    Soemthing about it maximising its correlation with Y, and how RSS is
    calculated off of YHat and thus PLS should produce more optimised Ys
    as it is directly optimised

## 3.4 --- Fisher's LDA / CDA (12)

We use `data/lda_data.csv`: four groups (`G1`--`G4`), 20 observations
per group, and 6 variables (`x1`--`x6`).

```{r}
lda_dat <- read.csv("data/lda_data.csv")
head(lda_dat)

dat_org = lda_dat
```

a.  **R code:** From first principles, code up the full Fisher's LDA
    algorithm: compute $\mathbf{B}$ and $\mathbf{W}$, obtain the
    eigendecomposition of $\mathbf{W}^{-1}\mathbf{B}$, and scale the
    eigenvectors so that
    $\mathbf{a}_k'\mathbf{S}_{pooled}\mathbf{a}_k = 1$. Display the
    resulting discriminant loadings. (6)

    ```{r}
    X = as.matrix(lda_dat[,-1])
    lda_dat$group = as.factor(lda_dat$group)
    g = length(unique(lda_dat$group))
    n_i = table(lda_dat$group)
    W = matrix(0, ncol(X), ncol(X))
    B = matrix(0, ncol(X), ncol(X))

    overall_means = colMeans(X)
    group_means = lda_dat %>% 
      group_by(group) %>% 
      summarise(across(everything(), mean), .groups = "drop") %>% 
      dplyr::select(-group) %>% 
      as.matrix()

    for(i in 1:g){
      diff = group_means[i,] - overall_means
      B = B + n_i[i] * (diff %*% t(diff))
      
      Xi = X[as.integer(lda_dat$group) == i,]
      W = W + cov(Xi) * (nrow(Xi) - 1)
    }

    eig_lda = eigen(solve(W) %*% B)
    vec = eig_lda$vectors %>% Re()
    Spooled = W / (sum(n_i) - g)
    sd = diag(sqrt(t(vec) %*% Spooled %*% vec))
    standardised = vec %*% diag(1/sd)

    kable(data.frame(Loading1 = standardised[,1], Loading2 = standardised[,2]), digits = 4)

    # b)
    lda_dat$LD1 = as.numeric(X %*% standardised[,1])
    lda_dat$LD2 = as.numeric(X %*% standardised[,2])
    ggplot(lda_dat, aes(x = LD1, y = LD2, col = group)) +
      geom_point()
    ```

    ```{r}
    ## CHECK ##
    X_org = as.matrix(dat_org[,-1])
    lda_fit = lda(group ~ ., data = dat_org)

    kable(data.frame(Loading1 = lda_fit$scaling[,1], Loading2 = lda_fit$scaling[,2]), digits = 4)

    dat_org$LD1 = X_org %*% lda_fit$scaling[,1]
    dat_org$LD2 = X_org %*% lda_fit$scaling[,2]

    ggplot(dat_org, aes(x = LD1, y = LD2, col = group)) +
      geom_point()
    ```

b.  **R code + plot + typed response:** Compute the discriminant scores
    and produce a scatterplot of the first two discriminants, coloured
    by group. Using the loadings, interpret the plot: state which
    discriminant separates which group(s), and which variables are
    driving that separation. (4)

    Overall G1 and G4 are seperated well while G2 and G3 are homogenous.
    We see that LD1 drives G4 to seperate from the other 3 groups. This
    seperation is driven strongly by x4 in the negative direction.
    pushing G4 away from the other groups. LD2 seperates G1 from G2
    while having the other groups somewhat in the middle. This
    seperation is driven by x1 which is in the positive direction

c.  **Typed response:** Fisher's LDA as coded in (a) will not always
    work. Describe, in general, the condition under which the algorithm
    breaks down, and why. (2)

    Since we have to fidn the inverse for W the algorithm will break
    when W is singular. This will occur when the Within variation
    produces a singular matrix. the circumstances underwhcih this would
    happen is if we have too few observations per group when compared to
    the total number of variables.

------------------------------------------------------------------------

# Multiple choice (4)

**Question A (1)** Suppose there is no correlation between any variables
from two sets of variables. If we applied canonical correlation
analysis, what would the values of the left canonical variate loadings
be for the first pair?

a.  Arbitrary, i.e. any vector in $\mathbb{R}^p$ that preserves unit
    variance
b.  0
c.  Impossible to say
d.  Proportional to a vector with $1/\sqrt{p}$ in every entry

**Question B (1)** Which scenario is likely to result in the largest
canonical correlation, all other things being equal?

a.  $X^{(1)}: 1 \times 1$ and $X^{(2)}: 1 \times 1$
b.  $X^{(1)}: 10 \times 1$ and $X^{(2)}: 1 \times 1$
c.  $X^{(1)}: 10 \times 1$ and $X^{(2)}: 10 \times 1$
d.  $X^{(1)}: 10 \times 1$ and $X^{(2)}: 100 \times 1$

**Question C (1)** Suppose we post-multiply the loadings matrix of an
orthogonal factor analysis by some orthonormal $\mathbf{Q}$. What
happens to the implied factors?

a.  Unchanged, as $\mathbf{Q}$ is orthonormal
b.  Unchanged, as the factors are never observed
c.  Pre-multiplied by $\mathbf{Q}'$ (i.e. become
    $\mathbf{Q}'\mathbf{F}$)
d.  Post-multiplied by $\mathbf{Q}'$ (i.e. become
    $\mathbf{F}\mathbf{Q}'$)

**Question D (1)** In Fisher's LDA, if the pooled within-group SSCP
matrix $\mathbf{W}$ is singular, which of the following is true?

a.  LDA can proceed exactly as usual since $\mathbf{B}$ is unaffected
b.  Some form of dimension reduction or regularisation is needed before
    $\mathbf{W}$ can be inverted
c.  $\mathbf{B}$ automatically becomes full rank to compensate
d.  The discriminant functions default to the first principal component
    only

**Question E (1)** A PCA scree plot is essentially flat (all eigenvalues
are roughly equal in size). What does this suggest?

a.  The variables are highly correlated, so a single PC will suffice
b.  The variables are close to uncorrelated with similar variances, so
    there is little redundancy for PCA to exploit and dimension
    reduction will not help much
c.  The data should not have been standardised before running PCA
d.  The first PC will still explain more than 90% of the total variance

------------------------------------------------------------------------
