---
title: "Canonical Correlation Analysis — Summary Sheet"
format: pdf
editor: visual
---

## What is CCA?

We have two groups of variables, $\mathbf{X}^{(1)}$ ($p \times 1$) and $\mathbf{X}^{(2)}$ ($q \times 1$). CCA finds linear combinations of each group --- called **canonical variates** --- that are maximally correlated with each other.

$$U_k = \mathbf{a}_k' \mathbf{X}^{(1)}, \qquad V_k = \mathbf{b}_k' \mathbf{X}^{(2)}$$

The pairs $(U_k, V_k)$ are found sequentially, each uncorrelated with all previous pairs. The correlation between $U_k$ and $V_k$ is the $k$-th **canonical correlation** $\rho_k^*$.

The maximum number of pairs is $\min(p, q)$, but in practice we only retain pairs with a meaningfully large $\rho_k^*$.

------------------------------------------------------------------------

## Partitioning the Covariance Matrix

The full covariance matrix is partitioned according to the two groups:

$$\boldsymbol{\Sigma} = \begin{pmatrix} \boldsymbol{\Sigma}_{11} & \boldsymbol{\Sigma}_{12} \\ \boldsymbol{\Sigma}_{21} & \boldsymbol{\Sigma}_{22} \end{pmatrix}$$

where $\boldsymbol{\Sigma}_{11}$ is the within-group covariance of $\mathbf{X}^{(1)}$, $\boldsymbol{\Sigma}_{22}$ is for $\mathbf{X}^{(2)}$, and $\boldsymbol{\Sigma}_{12} = \boldsymbol{\Sigma}_{21}'$ captures between-group covariances. If working with standardised variables, $\boldsymbol{\Sigma}$ is replaced by the correlation matrix $\boldsymbol{\rho}$.

For $U = \mathbf{a}'\mathbf{X}^{(1)}$ and $V = \mathbf{b}'\mathbf{X}^{(2)}$:

$$\text{Var}(U) = \mathbf{a}'\boldsymbol{\Sigma}_{11}\mathbf{a}, \qquad \text{Var}(V) = \mathbf{b}'\boldsymbol{\Sigma}_{22}\mathbf{b}, \qquad \text{Cov}(U,V) = \mathbf{a}'\boldsymbol{\Sigma}_{12}\mathbf{b}$$

We maximise:

$$\text{Corr}(U, V) = \frac{\mathbf{a}'\boldsymbol{\Sigma}_{12}\mathbf{b}}{\sqrt{\mathbf{a}'\boldsymbol{\Sigma}_{11}\mathbf{a} \cdot \mathbf{b}'\boldsymbol{\Sigma}_{22}\mathbf{b}}}$$

------------------------------------------------------------------------

## The Key Matrix R and Why $\boldsymbol{\Sigma}_{11}^{-1/2}$ Appears

To solve the maximisation problem we make the substitution $\mathbf{a}^* = \boldsymbol{\Sigma}_{11}^{1/2}\mathbf{a}$, so $\mathbf{a} = \boldsymbol{\Sigma}_{11}^{-1/2}\mathbf{a}^*$. This turns the unit variance constraint $\mathbf{a}'\boldsymbol{\Sigma}_{11}\mathbf{a} = 1$ into the simpler $\mathbf{a}^{*'}\mathbf{a}^* = 1$. The problem then reduces to a standard eigenvalue problem in $\mathbf{a}^*$, with solution given by the eigenvectors of:

$$\mathbf{R} = \boldsymbol{\Sigma}_{11}^{-1/2} \boldsymbol{\Sigma}_{12} \boldsymbol{\Sigma}_{22}^{-1} \boldsymbol{\Sigma}_{21} \boldsymbol{\Sigma}_{11}^{-1/2}$$

To get back the canonical coefficients in the original variable space we undo the substitution:

$$\mathbf{a}_k = \boldsymbol{\Sigma}_{11}^{-1/2} \mathbf{e}_k$$

where $\mathbf{e}_k$ is the $k$-th eigenvector of $\mathbf{R}$. So the $\boldsymbol{\Sigma}_{11}^{-1/2}$ rescales the eigenvector back into the original variable space, accounting for the variance structure of $\mathbf{X}^{(1)}$.

The squared canonical correlations $\rho_1^{*2} \geq \rho_2^{*2} \geq \cdots$ are the eigenvalues of $\mathbf{R}$.

------------------------------------------------------------------------

## Properties of Canonical Variates

$$\text{Var}(U_k) = \text{Var}(V_k) = 1$$

$$\text{Cov}(U_k, U_l) = \text{Cov}(V_k, V_l) = 0 \quad (k \neq l)$$

$$\text{Cov}(U_k, V_l) = 0 \quad (k \neq l)$$

$$\text{Corr}(U_k, V_k) = \rho_k^* \quad \text{(this is what we maximise)}$$

So within each group, different canonical variates are uncorrelated. Across groups, only matched pairs $(U_k, V_k)$ are correlated.

------------------------------------------------------------------------

## Canonical Loadings and Cross-Loadings

The **canonical coefficients** $\mathbf{a}_k$ and $\mathbf{b}_k$ are the weights on the original variables and should be avoided for interpretation due to multicollinearity.

Instead, use:

**Canonical loadings** --- correlation between each original variable and its own group's canonical variate:

$$\text{Corr}(X_j^{(1)}, U_k), \qquad \text{Corr}(X_j^{(2)}, V_k)$$

These tell you how much information $U_k$ (or $V_k$) captures about each original variable.

**Cross-loadings** --- correlation between original variables and the *other* group's canonical variate:

$$\text{Corr}(X_j^{(1)}, V_k), \qquad \text{Corr}(X_j^{(2)}, U_k)$$

These exist because $U_k$ and $V_k$ are maximally correlated, so information from one group bleeds into the other's variate. The stronger $\rho_k^*$, the more informative the cross-loadings.

**Signs matter** --- a positive loading means as the canonical variate increases, so does the original variable; negative means the opposite. If $U_1$ has positive loadings on $X_1^{(1)}$ and $X_2^{(1)}$ but negative cross-loadings on $X_1^{(2)}$ and $X_2^{(2)}$, this tells you the two groups move in opposite directions.

------------------------------------------------------------------------

## R Code: CCA from First Principles (Correlation Matrix)

```{r}
# Example correlation matrix (p = q = 2)
rho <- matrix(c(1.0, 0.4, 0.5, 0.6,
                0.4, 1.0, 0.3, 0.4,
                0.5, 0.3, 1.0, 0.2,
                0.6, 0.4, 0.2, 1.0), 4, 4, byrow = TRUE)

# Step 1: Partition
p <- 2; q <- 2
rho11 <- rho[1:p, 1:p]
rho12 <- rho[1:p, (p+1):(p+q)]
rho21 <- rho[(p+1):(p+q), 1:p]
rho22 <- rho[(p+1):(p+q), (p+1):(p+q)]

# Step 2: Compute rho11^{-1/2} via eigendecomposition
# We need the matrix square root of rho11^{-1}
# Achieved by: V %*% diag(sqrt(eigenvalues)) %*% V^{-1}
rho11inv <- solve(rho11)
eig11    <- eigen(rho11inv)
rho11inv.sqrt <- eig11$vectors %*% 
                 diag(sqrt(eig11$values)) %*% 
                 solve(eig11$vectors)

# Step 3: Compute rho22^{-1}
rho22inv <- solve(rho22)

# Step 4: Form R and obtain eigendecomposition
R     <- rho11inv.sqrt %*% rho12 %*% rho22inv %*% rho21 %*% rho11inv.sqrt
R_eig <- eigen(R)

# Step 5: Canonical correlations (square root of eigenvalues)
rho_star <- sqrt(R_eig$values)
rho_star
```

```{r}
# Step 6: Canonical coefficients for U (a_k = rho11^{-1/2} e_k)
# Sign flip optional — just for alignment with cc() output
A <- -1 * rho11inv.sqrt %*% R_eig$vectors
A  # columns are a_1, a_2, ...
```

```{r}
# Step 7: Unstandardised canonical coefficients for V
# b_k is proportional to rho22^{-1} rho21 a_k
B_p <- rho22inv %*% rho21 %*% A

# Step 8: Standardise B so that Var(V_k) = b_k' rho22 b_k = 1
# We need to divide each column of B_p by sqrt(b_k' rho22 b_k)
# This ensures the canonical variates V_k have unit variance
brhob <- sqrt(diag(t(B_p) %*% rho22 %*% B_p))
B <- B_p / matrix(brhob, nrow = nrow(B_p), ncol = ncol(B_p), byrow = TRUE)
B  # columns are b_1, b_2, ...
```

The standardisation in Step 8 is necessary because $\mathbf{b}_k \propto \boldsymbol{\rho}_{22}^{-1}\boldsymbol{\rho}_{21}\mathbf{a}_k$ only guarantees the correct direction, not unit variance. We scale each column by $1/\sqrt{\mathbf{b}_k'\boldsymbol{\rho}_{22}\mathbf{b}_k}$ so that $\text{Var}(V_k) = \mathbf{b}_k'\boldsymbol{\rho}_{22}\mathbf{b}_k = 1$.

```{r}
# Step 9: Canonical loadings
# Corr(X^(1), U_k) = A %*% rho11
loadings_U <- t(A) %*% rho11
loadings_U

# Corr(X^(2), V_k) = B %*% rho22
loadings_V <- t(B) %*% rho22
loadings_V
```

```{r}
# Step 10: Cross-loadings
# Corr(X^(1), V_k) = A %*% rho12
cross_UV <- t(A) %*% rho12
cross_UV

# Corr(X^(2), U_k) = B %*% rho21
cross_VU <- t(B) %*% rho21
cross_VU
```

------------------------------------------------------------------------

## R Code: CCA Using the `CCA` Package

```{r}
#| eval: false
library(CCA)

# cc() takes the raw data matrices for each group
cc1 <- cc(X1, X2)

# Canonical correlations
cc1$cor

# Raw canonical coefficients
cc1$xcoef  # a_k coefficients
cc1$ycoef  # b_k coefficients

# Standardised coefficients (multiply by diagonal of SDs)
s1 <- diag(sqrt(diag(cov(X1))))
s2 <- diag(sqrt(diag(cov(X2))))
s1 %*% cc1$xcoef
s2 %*% cc1$ycoef

# Canonical variate scores for each observation
cc1$scores$xscores  # U_k scores
cc1$scores$yscores  # V_k scores

# Canonical loadings
cc1$scores$corr.X.xscores  # Corr(X^(1), U_k)
cc1$scores$corr.Y.yscores  # Corr(X^(2), V_k)

# Cross-loadings
cc1$scores$corr.X.yscores  # Corr(X^(1), V_k)
cc1$scores$corr.Y.xscores  # Corr(X^(2), U_k)
```

------------------------------------------------------------------------

## Interpretation Summary

| Quantity                       | What it tells you                                                                     |
|--------------------------------|---------------------------------------------------------------------------------------|
| $\rho_k^*$                     | How correlated the $k$-th pair of canonical variates are                              |
| $\mathbf{a}_k$, $\mathbf{b}_k$ | Weights forming $U_k$ and $V_k$ --- avoid for interpretation due to multicollinearity |
| $\text{Corr}(X_j^{(1)}, U_k)$  | How much information $U_k$ captures about $X_j^{(1)}$                                 |
| $\text{Corr}(X_j^{(2)}, V_k)$  | How much information $V_k$ captures about $X_j^{(2)}$                                 |
| $\text{Corr}(X_j^{(1)}, V_k)$  | Cross-loading: how much $X_j^{(1)}$ bleeds into $V_k$ via the canonical correlation   |
| Sign of loading                | Positive = move together; Negative = move in opposite directions                      |
