---
title: "Factor Analysis Summary"
format: pdf
editor: visual
---

```{r setup, include=FALSE}
library(knitr)
# Load air pollution data
if (!requireNamespace("DataTidy23RodoSTA2005SAssignment", quietly = TRUE)) {
  if (!requireNamespace("remotes", quietly = TRUE)) {
    utils::install.packages("remotes")
  }
  remotes::install_github("MiguelRodo/DataTidy23RodoSTA2005SAssignment")
}
if (!requireNamespace("tibble", quietly = TRUE)) {
  install.packages("tibble")
}
library(tibble) 
library(knitr)
data("data_tidy_yield", package = "DataTidyRodoSTA2005S")

library(readr)
dat2 = read_csv("../../../../Section_B/Lecture_9/CA9/Air Pollution Data .csv")
ap <- dat2
```

# The Orthogonal Factor Model

The orthogonal factor model (OFM) decomposes observed variables $\mathbf{X}$ as:

$$\mathbf{X} - \boldsymbol{\mu} = \mathbf{L}\mathbf{F} + \boldsymbol{\epsilon}$$

where $\mathbf{X}$ is $p \times 1$, $\mathbf{F}$ is $m \times 1$ (unobserved common factors), $\boldsymbol{\epsilon}$ is $p \times 1$ (specific factors), and $\mathbf{L}$ is the $p \times m$ loadings matrix.

**Assumptions:**

-   $E[\mathbf{F}] = \mathbf{0}$, $\text{Var}[\mathbf{F}] = \mathbf{I}$
-   $E[\boldsymbol{\epsilon}] = \mathbf{0}$, $\text{Var}[\boldsymbol{\epsilon}] = \boldsymbol{\Psi}$ (diagonal)
-   $\text{Cov}[\mathbf{F}, \boldsymbol{\epsilon}] = \mathbf{0}$

The implied covariance structure is:

$$\text{Cov}[\mathbf{X}] = \mathbf{L}\mathbf{L}' + \boldsymbol{\Psi}$$

with $\text{Var}(X_i) = \sum_{j=1}^m l_{ij}^2 + \psi_{ii}$ and $\text{Cov}(X_i, X_j) = \sum_{k=1}^m l_{ik}l_{jk}$.

The covariance between observed and unobserved variables is $\text{Cov}[\mathbf{X}, \mathbf{F}] = \mathbf{L}$.

# Communality and Specific Variance

Working on the correlation matrix (standardised data), $\text{Var}(X_i) = 1$, so:

$$1 = h_i^2 + \psi_{ii}$$

where $h_i^2 = \sum_{j=1}^m l_{ij}^2$ is the **communality** (variance explained by the common factors) and $\psi_{ii} = 1 - h_i^2$ is the **specific variance**.

# Non-Uniqueness and Rotation

For any $m \times m$ orthonormal matrix $\mathbf{T}$ (i.e. $\mathbf{T}\mathbf{T}' = \mathbf{I}$), the rotated solution $\mathbf{L}^* = \mathbf{L}\mathbf{T}$ is equivalent because:

$$\mathbf{L}^*\mathbf{L}^{*'} + \boldsymbol{\Psi} = \mathbf{L}\mathbf{T}\mathbf{T}'\mathbf{L}' + \boldsymbol{\Psi} = \mathbf{L}\mathbf{L}' + \boldsymbol{\Psi}$$

This means we can rotate the loading matrix to improve interpretability --- seeking a solution where each factor loads strongly on a few variables and weakly on the rest --- without changing the model fit or $\boldsymbol{\Psi}$.

# Method 1: Principal Component Method

The PC method approximates $\mathbf{L}$ using the eigendecomposition of the sample correlation matrix $\mathbf{R} = \mathbf{V}\boldsymbol{\Lambda}\mathbf{V}'$.

The loadings for $m$ factors are:

$$\hat{\mathbf{L}} = \mathbf{V}_m \boldsymbol{\Lambda}_m^{1/2}$$

where $\mathbf{V}_m$ contains the top $m$ eigenvectors and $\boldsymbol{\Lambda}_m = \text{diag}(\lambda_1, \ldots, \lambda_m)$.

The communalities and specific variances are then:

$$\hat{h}_i^2 = \sum_{j=1}^m \hat{l}_{ij}^2, \qquad \hat{\psi}_{ii} = 1 - \hat{h}_i^2$$

```{r pc-method}
# Correlation matrix
R <- cor(ap)

# Eigendecomposition
eig <- eigen(R)
p <- ncol(ap)
m <- 2  # number of factors

# Loadings: p x m
V_m <- eig$vectors[, 1:m]
Lambda_m <- diag(sqrt(eig$values[1:m]))
L_pc <- V_m %*% Lambda_m
rownames(L_pc) <- colnames(ap)
colnames(L_pc) <- paste0("Factor", 1:m)

# Communalities and specific variances
h2_pc <- rowSums(L_pc^2)
psi_pc <- 1 - h2_pc

# Reconstructed correlation matrix
R_hat_pc <- L_pc %*% t(L_pc) + diag(psi_pc)

kable(
  data.frame(L_pc, Communality = h2_pc, Specific = psi_pc),
  digits = 3,
  caption = "PC Method: Loadings, Communalities, and Specific Variances"
)
```

```{r pc-reconstructed}
kable(
  round(R - R_hat_pc, 3),
  caption = "Residual: Sample minus PC-implied Correlation Matrix"
)
```

# Method 2: Maximum Likelihood

ML factor analysis assumes $\mathbf{X} \sim N_p(\boldsymbol{\mu}, \mathbf{L}\mathbf{L}' + \boldsymbol{\Psi})$ and finds $\hat{\mathbf{L}}$ and $\hat{\boldsymbol{\Psi}}$ that maximise the likelihood. This allows formal likelihood ratio testing of the number of factors.

```{r ml-method}
FA <- factanal(ap, factors = m, rotation = "none")

L_ml <- as.matrix(FA$loadings[, 1:m])
colnames(L_ml) <- paste0("Factor", 1:m)
uniq <- FA$uniquenesses
h2_ml <- 1 - uniq

kable(
  data.frame(L_ml, Communality = h2_ml, Specific = uniq),
  digits = 3,
  caption = "ML Method: Loadings, Communalities, and Specific Variances"
)
```

```{r ml-reconstructed}
R_hat_ml <- L_ml %*% t(L_ml) + diag(uniq)
kable(
  round(R - R_hat_ml, 3),
  caption = "Residual: Sample minus ML-implied Correlation Matrix"
)
```

# Comparison of PC and ML Methods

```{r comparison}
df_compare <- data.frame(
  PC_F1      = L_pc[, 1],
  PC_F2      = L_pc[, 2],
  PC_h2      = h2_pc,
  ML_F1      = L_ml[, 1],
  ML_F2      = L_ml[, 2],
  ML_h2      = h2_ml
)
kable(df_compare, digits = 3,
      caption = "Comparison of PC and ML Loadings and Communalities")
```

The PC method greedily maximises variance explained sequentially, tending to produce higher communalities. ML is more conservative, penalising poor-fitting variables with higher specific variances.

# Varimax Rotation

Varimax rotation finds the orthonormal $\mathbf{T}$ that maximises the variance of the squared loadings within each factor, pushing each factor toward having a few large and many near-zero loadings.

```{r rotation}
# Rotate PC loadings
L_pc_rot <- varimax(L_pc)$loadings
L_pc_rot <- as.matrix(L_pc_rot)
colnames(L_pc_rot) <- paste0("Factor", 1:m)

# Rotate ML loadings
FA_rot <- factanal(ap, factors = m, rotation = "varimax")
L_ml_rot <- as.matrix(FA_rot$loadings[, 1:m])
colnames(L_ml_rot) <- paste0("Factor", 1:m)

kable(
  data.frame(L_pc_rot, row.names = colnames(ap)),
  digits = 3,
  caption = "Varimax-Rotated PC Loadings"
)

kable(
  data.frame(L_ml_rot, row.names = colnames(ap)),
  digits = 3,
  caption = "Varimax-Rotated ML Loadings"
)
```

Note that communalities are unchanged by rotation: $h_i^2 = \sum_j l_{ij}^{*2} = \sum_j l_{ij}^2$.

# Factor Scores (WLS)

Given estimated loadings $\hat{\mathbf{L}}$ and specific variances $\hat{\boldsymbol{\Psi}}$, the weighted least squares factor scores for a mean-centred observation are:

$$\hat{\mathbf{F}} = (\hat{\mathbf{L}}'\hat{\boldsymbol{\Psi}}^{-1}\hat{\mathbf{L}})^{-1}\hat{\mathbf{L}}'\hat{\boldsymbol{\Psi}}^{-1}(\mathbf{x} - \bar{\mathbf{x}})$$

Variables with small specific variance get more weight as they are more reliably driven by the common factors.

```{r factor-scores}
# Refit on raw data for score computation
FA_raw <- factanal(ap, factors = m, rotation = "none")
L_fs   <- as.matrix(FA_raw$loadings[, 1:m])
Psi_inv <- diag(1 / FA_raw$uniquenesses)

# WLS coefficient matrix
A      <- t(L_fs) %*% Psi_inv %*% L_fs
B      <- solve(A) %*% t(L_fs) %*% Psi_inv

# Standardise data and compute scores
Z      <- scale(ap)
scores <- Z %*% t(B)
colnames(scores) <- paste0("Factor", 1:m)

kable(
  head(round(scores, 3), 10),
  caption = "WLS Factor Scores (ML, first 10 observations)"
)
```

```{r factor-scores-plot}
plot(scores[, 1], scores[, 2],
     xlab = "Factor 1", ylab = "Factor 2",
     main = "Factor Scores: Factor 1 vs Factor 2",
     pch = 19, col = "steelblue")
abline(h = 0, v = 0, col = "gray60", lty = 2)
```

A random scatter with no visible pattern in this plot indicates the factors are approximately uncorrelated, as assumed by the OFM.
