---
title: "Least-Squares Regression: Multiple, Multivariate, PCR and PLS"
format: pdf
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(knitr)
library(pls)
data("data_tidy_house_price", package = "DataTidy23RodoHonsMult")
data("data_tidy_paper", package = "DataTidy23RodoHonsMult")
```

# Overview

This document covers four regression approaches:

-   **Simple/Multiple regression (MLR)**: one response, one or more
    predictors
-   **Multivariate regression (MVR)**: multiple responses, multiple
    predictors
-   **Principal Components Regression (PCR)**: MLR on PCA components of
    predictors
-   **Partial Least Squares (PLS)**: regression on components that
    maximise covariance with response

------------------------------------------------------------------------

# Notation

| Symbol | Meaning |
|----------------------------------|--------------------------------------|
| $n$ | number of observations |
| $r$ | number of responses ($r = 1$ for MLR, $r \geq 1$ for MVR) |
| $k$ | number of predictors **including** the intercept |
| $p = k - 1$ | number of predictors **excluding** the intercept |
| $\mathbf{y} : n \times 1$ | response vector (MLR) |
| $\mathcal{Y} : n \times r$ | response matrix (MVR) |
| $\mathbf{X} : n \times k$ | design matrix (first column all 1s for intercept) |
| $\boldsymbol{\beta} : k \times 1$ | coefficient vector (MLR) |
| $\mathcal{B} : k \times r$ | coefficient matrix (MVR) |

------------------------------------------------------------------------

# 1. Multiple Regression (MLR)

## Model

One response, $p$ predictors:

$$Y = \beta_0 + \beta_1 X_1 + \ldots + \beta_p X_p + \epsilon, \quad \text{or in matrix form} \quad \mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}$$

## Assumptions

-   **Linearity**:
    $\mathrm{E}[Y] = \beta_0 + \beta_1 X_1 + \ldots + \beta_p X_p$
-   **Constant variance**: $\mathrm{Var}[\epsilon] = \sigma^2$
-   **Independence**: $\mathrm{Cov}[\epsilon_i, \epsilon_j] = 0$ for
    $i \neq j$ — errors across *different observations* are uncorrelated

## Estimating Coefficients

We choose $\hat{\boldsymbol{\beta}}$ to minimise the sum of squared
residuals:

$$\sum_{i=1}^n (y_i - \hat{y}_i)^2 = (\mathbf{y} - \mathbf{X}\hat{\boldsymbol{\beta}})'(\mathbf{y} - \mathbf{X}\hat{\boldsymbol{\beta}})$$

This gives the OLS estimator:

$$\hat{\boldsymbol{\beta}} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{y}$$

## The Hat Matrix

Fitted values for all observations at once:

$$\hat{\mathbf{y}} = \mathbf{X}\hat{\boldsymbol{\beta}} = \mathbf{X}(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{y} = \mathbf{H}\mathbf{y}$$

where $\mathbf{H} = \mathbf{X}(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'$
is called the **hat matrix**. It is:

-   A projection matrix:
    $\mathbf{H}\mathbf{y} = \mathrm{proj}_{\mathbf{X}}(\mathbf{y})$
-   Idempotent: $\mathbf{H}^2 = \mathbf{H}$

## Estimator Properties

$$\mathrm{E}[\hat{\boldsymbol{\beta}}] = \boldsymbol{\beta} \quad \text{(unbiased)}$$
$$\mathrm{Cov}[\hat{\boldsymbol{\beta}}] = \sigma^2(\mathbf{X}'\mathbf{X})^{-1}$$
$$\mathrm{Cov}[\hat{\boldsymbol{\epsilon}}] = \sigma^2(\mathbf{I} - \mathbf{H})$$

Note: the true errors $\epsilon_i$ are independent, but the
**residuals** $\hat{\epsilon}_i$ are correlated with each other because
they all depend on the same $\hat{\boldsymbol{\beta}}$. The off-diagonal
entries of $\sigma^2(\mathbf{I} - \mathbf{H})$ capture this.

## Estimating the Variance

$$\hat{\sigma}^2 = s^2 = \frac{1}{n-k}(\mathbf{y} - \mathbf{X}\hat{\boldsymbol{\beta}})'(\mathbf{y} - \mathbf{X}\hat{\boldsymbol{\beta}})$$

We divide by $n - k$ (not $n$) because we lose one degree of freedom for
each of the $k$ parameters estimated. This makes $s^2$ an unbiased
estimator of $\sigma^2$.

## Sums of Squares

$$\mathrm{SST} = \sum_{i=1}^n (y_i - \bar{y})^2 = \mathbf{y}'\mathbf{y} - n\bar{y}^2$$

$$\mathrm{SST} = \mathrm{SSR} + \mathrm{SSE}$$

where SSR is the sum of squares due to regression and SSE the sum of
squares due to error:

$$\mathrm{SSR} = \hat{\mathbf{y}}'\hat{\mathbf{y}} - n\bar{y}^2, \qquad \mathrm{SSE} = \mathbf{y}'\mathbf{y} - \hat{\mathbf{y}}'\hat{\mathbf{y}}$$

$$R^2 = \frac{\mathrm{SSR}}{\mathrm{SST}} = 1 - \frac{\mathrm{SSE}}{\mathrm{SST}}$$

$R^2$ is the proportion of total variance explained by the model,
$0 \leq R^2 \leq 1$.

## Inference (requires normality assumption)

Assuming
$\boldsymbol{\epsilon} \sim \mathcal{N}(\mathbf{0}, \sigma^2 \mathbf{I})$,
it follows that:

$$\hat{\boldsymbol{\beta}} \sim \mathcal{N}\left(\boldsymbol{\beta},\ \sigma^2(\mathbf{X}'\mathbf{X})^{-1}\right), \qquad \frac{n-k}{\sigma^2}s^2 \sim \chi^2_{n-k}$$

Note: $\hat{\boldsymbol{\beta}}$ and $s^2$ are independent, because they
are functions of orthogonal projections of $\mathbf{y}$ —
$\hat{\boldsymbol{\beta}}$ depends on $\mathbf{H}\mathbf{y}$ and $s^2$
depends on $(\mathbf{I} - \mathbf{H})\mathbf{y}$, and orthogonal
projections of a normal vector are independent.

### Confidence Interval for $\beta_j$

$$\hat{\beta}_j \pm t_{n-k}^{\alpha/2} \sqrt{s^2 (\mathbf{X}'\mathbf{X})^{-1}_{jj}}$$

### Univariate Hypothesis Test: $H_0 : \beta_j = 0$

$$t = \frac{\hat{\beta}_j}{\sqrt{s^2(\mathbf{X}'\mathbf{X})^{-1}_{jj}}}, \qquad \text{reject if } |t| > t_{n-k}^{1-\alpha/2}$$

### Simultaneous Hypothesis Test: $H_0 : \boldsymbol{\beta}^{(1)} = \mathbf{0}$

Where $\boldsymbol{\beta}^{(1)} : q \times 1$ is a subset of
coefficients and
$(\mathbf{X}'\mathbf{X})^{-1} = \mathbf{C} = \begin{bmatrix} \mathbf{C}_{11} & \mathbf{C}_{12} \\ \mathbf{C}_{21} & \mathbf{C}_{22} \end{bmatrix}$:

$$F = \frac{\hat{\boldsymbol{\beta}}^{(1)\prime}\mathbf{C}_{11}^{-1}\hat{\boldsymbol{\beta}}^{(1)}}{qs^2}, \qquad \text{reject if } F \geq F^{1-\alpha}_{q,\, n-k}$$

------------------------------------------------------------------------

## MLR Example: House Prices

We model selling price against house size and assessed value.

```{r mlr-data}
# Extract response vector and design matrix
y_vec <- data_tidy_house_price %>% pull(selling_price)

X_mat <- data_tidy_house_price %>%
  select(-selling_price) %>%
  as.matrix() %>%
  cbind(intercept = 1, .)
```

```{r mlr-coefficients}
# Estimate coefficients from scratch: beta = (X'X)^-1 X'y
beta_vec <- solve(t(X_mat) %*% X_mat) %*% t(X_mat) %*% y_vec
colnames(beta_vec) <- "estimate"
beta_vec %>% signif(3) %>% kable(caption = "MLR coefficient estimates")
```

```{r mlr-r2}
# Calculate R^2
SST <- sum((y_vec - mean(y_vec))^2)
SSE <- sum((y_vec - X_mat %*% beta_vec)^2)
SSR <- SST - SSE
R2  <- SSR / SST
cat("R^2 =", signif(R2, 2))
```

```{r mlr-verify}
# Verify with lm()
lm(selling_price ~ ., data = data_tidy_house_price) %>%
  coef() %>%
  signif(3) %>%
  kable(caption = "lm() output for comparison")
```

------------------------------------------------------------------------

# 2. Multivariate Regression (MVR)

## What is MVR?

MVR is simply running MLR for each of $r$ response variables
simultaneously, using the same predictor matrix $\mathbf{X}$ for all of
them.

## Model

$$\mathcal{Y} = \mathbf{X}\mathcal{B} + \mathcal{E}$$

where $\mathcal{Y} : n \times r$, $\mathcal{B} : k \times r$,
$\mathcal{E} : n \times r$.

Fitted value for observation $i$:
$\hat{\mathcal{Y}}_{[i]} = \mathbf{x}_i'\hat{\mathcal{B}}$ (a
$1 \times r$ row vector).

## Assumptions

-   **Linearity**: $\mathrm{E}[\mathcal{Y}] = \mathbf{X}\mathcal{B}$
-   **Constant variance**:
    $\mathrm{Var}[\mathcal{E}] = \boldsymbol{\Sigma}$ — an $r \times r$
    matrix where the diagonal entries are the individual response
    variances and the off-diagonals are the covariances *between
    responses within the same observation*. This does **not** have to be
    zero.
-   **Independence**:
    $\mathrm{Cov}[\mathcal{E}_i, \mathcal{E}_j] = \mathbf{0}$ for
    $i \neq j$ — errors across *different observations* must be
    independent. Different responses on the same observation can be
    correlated; different observations cannot.

## Estimating Coefficients

$$\hat{\mathcal{B}} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathcal{Y}$$

Each column of $\hat{\mathcal{B}}$ is exactly the OLS estimate you would
get running a separate MLR for that response. $\boldsymbol{\Sigma}$ does
not affect coefficient estimation — only inference.

## Estimating the Variance-Covariance Matrix

$$\hat{\boldsymbol{\Sigma}} = \mathbf{S} = \frac{1}{n-k}(\mathcal{Y} - \mathbf{X}\hat{\mathcal{B}})'(\mathcal{Y} - \mathbf{X}\hat{\mathcal{B}})$$

Same $n - k$ denominator as MLR — we still lose $k$ degrees of freedom
for the $k$ rows of $\hat{\mathcal{B}}$.

## MVR Example: Paper Properties

We model four paper properties against four wood-pulp predictors.

```{r mvr-data}
# Response matrix: all paper columns
Y_mat <- data_tidy_paper %>%
  select(starts_with("paper")) %>%
  as.matrix()

# Design matrix: intercept + all pulp columns
X_mat <- data_tidy_paper %>%
  select(starts_with("pulp")) %>%
  as.matrix() %>%
  cbind(intercept = 1, .)
```

```{r mvr-coefficients}
# Estimate B: (X'X)^-1 X'Y
B_mat <- solve(t(X_mat) %*% X_mat) %*% t(X_mat) %*% Y_mat
B_mat %>% signif(3) %>% kable(caption = "MVR coefficient estimates")
```

```{r mvr-verify}
# Verify with lm()
lm(cbind(paper_1, paper_2, paper_3, paper_4) ~ pulp_1 + pulp_2 + pulp_3 + pulp_4,
   data = data_tidy_paper) %>%
  coef() %>%
  signif(3) %>%
  kable(caption = "lm() output for comparison")
```

------------------------------------------------------------------------

## Inference in MVR

When testing hypotheses *within* a single response variable, MVR and MLR
give exactly the same answer. The extra power of MVR is that you can
test hypotheses *across* response variables simultaneously.

### Individual Coefficient CI and t-test (same as MLR)

```{r mvr-inference-setup}
# Using only paper_1 as response for the inference example
Y_mat_1 <- data_tidy_paper[, "paper_1", drop = FALSE] %>% as.matrix()
colnames(Y_mat_1) <- "paper_1"

beta_vec <- solve(t(X_mat) %*% X_mat) %*% t(X_mat) %*% Y_mat_1
colnames(beta_vec) <- "estimate"

# Variance estimate
n <- nrow(Y_mat_1)
k <- ncol(X_mat)
var_resp <- sum((Y_mat_1 - X_mat %*% beta_vec)^2) / (n - k)

# Standard error for pulp_1 coefficient (row 2)
XtX_inv <- solve(t(X_mat) %*% X_mat)
se_pulp1 <- sqrt(var_resp * XtX_inv[2, 2])
```

```{r mvr-ci}
# 95% CI for beta_pulp1
t_crit <- qt(0.975, df = n - k)
ci <- beta_vec[2] + c(-1, 1) * t_crit * se_pulp1
cat("95% CI for beta_pulp1: [", signif(ci[1], 3), ",", signif(ci[2], 3), "]")
```

```{r mvr-ttest}
# t-test: H0: beta_pulp1 = 0
t_stat  <- beta_vec[2] / se_pulp1
p_value <- 2 * pt(-abs(t_stat), df = n - k)
cat("t =", signif(t_stat, 2), "  p-value =", signif(p_value, 2))
```

### Simultaneous F-test: $H_0 : \boldsymbol{\beta}^{(1)} = \mathbf{0}$

Testing
$H_0: \beta_{\text{pulp}_1} = \beta_{\text{pulp}_3} = \beta_{\text{pulp}_4} = 0$.

```{r mvr-ftest}
# Extract C11 block (rows/cols for pulp_1, pulp_3, pulp_4 = indices 2, 4, 5)
C11_mat <- XtX_inv[c(2, 4, 5), c(2, 4, 5)]

# Extract the corresponding beta estimates
beta_vec_1 <- beta_vec[c(2, 4, 5), , drop = FALSE]
q <- length(beta_vec_1)

# F statistic
f_stat <- (t(beta_vec_1) %*% solve(C11_mat) %*% beta_vec_1) / (q * var_resp)
p_value_f <- pf(f_stat, df1 = q, df2 = n - k, lower.tail = FALSE)

cat("F =", signif(f_stat, 2), "  p-value =", signif(p_value_f, 2))
```

------------------------------------------------------------------------

### Multivariate Likelihood Ratio Test: $H_0 : \mathcal{B}^{(1)} = \mathbf{0}$

This tests whether a subset of predictors has *any* effect on *any*
response. We compare a full model (all predictors) against a reduced
model (predictors under $H_0$ removed).

The test statistic is the likelihood ratio:

$$\Lambda = \left(\frac{|\hat{\boldsymbol{\Sigma}}|}{|\hat{\boldsymbol{\Sigma}}^{(2)}|}\right)^{n/2}$$

For large $n$, the following approximation holds:

$$-[n - k - 0.5(r - p + q + 1)]\ln(\Lambda^{2/n}) \sim \chi^2_{rq}$$

```{r mvr-lr-test}
# Test whether pulp_2 and pulp_3 have any effect on any paper property
# Full model uses all pulps; reduced model removes pulp_2 and pulp_3 (cols 3:4)
X_mat_2 <- X_mat[, -(3:4), drop = FALSE]

B_mat_full    <- solve(t(X_mat)   %*% X_mat)   %*% t(X_mat)   %*% Y_mat
B_mat_reduced <- solve(t(X_mat_2) %*% X_mat_2) %*% t(X_mat_2) %*% Y_mat

# MLE covariance matrices (divide by n, not n-k, for the LR test)
n <- nrow(Y_mat)
Sigma_full    <- t(Y_mat - X_mat   %*% B_mat_full)    %*% (Y_mat - X_mat   %*% B_mat_full)    / n
Sigma_reduced <- t(Y_mat - X_mat_2 %*% B_mat_reduced) %*% (Y_mat - X_mat_2 %*% B_mat_reduced) / n

r <- ncol(Y_mat)
k <- ncol(X_mat)
p <- k - 1
q <- ncol(X_mat) - ncol(X_mat_2)   # number of predictors being tested

test_stat <- -(n - k - 0.5 * (r - p + q + 1)) * log(det(Sigma_full) / det(Sigma_reduced))
p_value_lr <- pchisq(test_stat, df = r * q, lower.tail = FALSE)

# Form 1: compute Lambda first, then the test statistic / can also do this
Lambda   <- (det(Sigma_full) / det(Sigma_reduced))^(n / 2)
test_stat <- -(n - k - 0.5 * (r - p + q + 1)) * log(Lambda^(2 / n))

cat("Chi-sq =", signif(test_stat, 2), "  df =", r * q, "  p-value =", signif(p_value_lr, 3))
```

------------------------------------------------------------------------

# 3. Principal Components Regression (PCR)

## Idea

Replace the original $p$ predictors with $m \leq p$ principal
components, then run MLR on those components. Motivations:

-   Avoid multicollinearity (components are uncorrelated by
    construction)
-   Reduce overfitting (fewer predictors)
-   Improve interpretability

**Key point**: PCA is performed only on $\mathbf{X}$ — it ignores
$\mathbf{y}$ completely. The components maximise variance in the
predictor space, not association with the response.

## Method

1.  Compute the principal components of the (non-intercept) predictors
    (standardise first).
2.  Regress $\mathbf{y}$ on the first $m$ components.
3.  Choose $m$ by cross-validation or scree plot.

When cross-validating, PCA loadings must be **recomputed within each
fold** — they cannot be computed on the full dataset first.

```{r pcr-fit}
# Fit PCR using plsr() with cross-validation to choose m
# ncomp sets the maximum number of components to try
pcr_fit <- pcr(
  paper_1 ~ pulp_1 + pulp_2 + pulp_3 + pulp_4,
  data      = data_tidy_paper,
  scale     = TRUE,
  ncomp     = 4,
  validation = "CV"
)

# Extract cross-validated MSEP manually (avoids validationplot legend bug)
msep_vals <- MSEP(pcr_fit)$val[1, 1, ]   # [estimator, response, ncomp]
ncomp_seq <- 0:(length(msep_vals) - 1)

tibble(ncomp = ncomp_seq, MSEP = msep_vals) %>%
  ggplot(aes(x = ncomp, y = MSEP)) +
  geom_line() +
  geom_point() +
  labs(title = "PCR: Cross-validated MSEP", x = "Number of components", y = "CV-MSEP") +
  theme_minimal()
```

```{r pcr-optimal}
# Choose optimal number of components (minimum CV-MSEP)
m_opt <- which.min(msep_vals) - 1   # subtract 1 because index 1 = 0 components
cat("Optimal number of PCR components:", m_opt)
```

```{r pcr-coefficients}
# Coefficients in original variable space for chosen m
coef(pcr_fit, ncomp = m_opt, intercept = TRUE) %>%
  kable(caption = paste("PCR coefficients (m =", m_opt, ")"))
```

------------------------------------------------------------------------

# 4. Partial Least Squares (PLS)

## Idea

Like PCR, PLS takes linear combinations of the predictors and regresses
on them. The difference is *how* the components are chosen:

| Method | Components maximise                                        |
|--------|------------------------------------------------------------|
| PCR    | Variance in $\mathbf{X}$ (ignores $\mathbf{y}$)            |
| PLS    | Covariance between $\mathbf{X}\mathbf{a}$ and $\mathbf{y}$ |

This means PLS components are directly guided by the response — a
component that explains a lot of predictor variance but has nothing to
do with $\mathbf{y}$ will not be selected.

## Orthogonality

In PCA, the **loadings** $\mathbf{a}_i$ are orthogonal, which means the
**scores** $\mathbf{t}_i = \mathbf{X}\mathbf{a}_i$ are also orthogonal
as a consequence.

In PLS, the **loadings** are not orthogonal. Instead it is the
**scores** that are forced to be orthogonal, achieved through a
**deflation step** — after each component, its contribution is
subtracted from $\mathbf{X}$ before the next component is found. This is
why back-transforming PLS coefficients to the original variable space
requires extra work.

## Fitting with `plsr()`

```{r pls-fit}
pls_fit <- plsr(
  paper_1 ~ pulp_1 + pulp_2 + pulp_3 + pulp_4,
  data       = data_tidy_paper,
  scale      = TRUE,
  ncomp      = 4,
  validation = "CV"
)

# Extract CV-MSEP manually
msep_pls <- MSEP(pls_fit)$val[1, 1, ]
ncomp_seq <- 0:(length(msep_pls) - 1)

tibble(ncomp = ncomp_seq, MSEP = msep_pls) %>%
  ggplot(aes(x = ncomp, y = MSEP)) +
  geom_line() +
  geom_point() +
  labs(title = "PLS: Cross-validated MSEP", x = "Number of components", y = "CV-MSEP") +
  theme_minimal()
```

```{r pls-optimal}
m_opt_pls <- which.min(msep_pls) - 1
cat("Optimal number of PLS components:", m_opt_pls)
```

```{r pls-coefficients}
# Coefficients back-transformed to original variable space
coef(pls_fit, ncomp = m_opt_pls, intercept = TRUE) %>%
  kable(caption = paste("PLS coefficients (m =", m_opt_pls, ")"))
```

------------------------------------------------------------------------

# PCR vs PLS: A Comparison

```{r pcr-vs-pls}
# Plot CV-MSEP for both methods side by side
bind_rows(
  tibble(ncomp = 0:(length(msep_vals) - 1), MSEP = msep_vals, method = "PCR"),
  tibble(ncomp = 0:(length(msep_pls)  - 1), MSEP = msep_pls,  method = "PLS")
) %>%
  ggplot(aes(x = ncomp, y = MSEP, colour = method)) +
  geom_line() +
  geom_point() +
  labs(title = "PCR vs PLS: Cross-validated MSEP",
       x = "Number of components", y = "CV-MSEP", colour = "Method") +
  theme_minimal()
```

PLS typically needs fewer components than PCR to achieve the same
prediction error, because its components are already oriented toward the
response.
