---
title: "PCA – Code Walkthrough"
format: pdf
editor: 
  markdown: 
    wrap: 72
---

# PCA Code Walkthrough

## 1. Loading the data

```{r}
#| message: false
if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
remotes::install_github("MiguelRodo/DataTidy23RodoHonsMult@2024", quiet = TRUE)

data("data_tidy_turtle", package = "DataTidy23RodoHonsMult")
head(data_tidy_turtle)
```

49 observations with `gender`, `length`, `width`, and `height`. The
measurements are on similar scales and highly correlated, making PCA
appropriate. We log-transform to linearise relationships between
morphological measurements.

------------------------------------------------------------------------

## 2. PCA using `prcomp`

```{r}
pr_obj <- prcomp(
  ~ log(length) + log(width) + log(height),
  data = data_tidy_turtle,
  retx = TRUE
)
pr_obj
```

-   The formula specifies which variables to include and applies
    `log()`.
-   `retx = TRUE` returns the PC scores, stored in `pr_obj$x`.
-   Under the hood, `prcomp` centres the data and applies the SVD.

The output shows:

-   **Standard deviations** — $\sqrt{\lambda_k}$ for each component.
-   **Rotation** — the loading matrix $\mathbf{V}$; each column is a
    loading vector $\mathbf{a}_k$.

PC1 has roughly equal loadings across all three variables, so it acts as
a general size index: larger turtles score higher on PC1.

------------------------------------------------------------------------

## 3. Inspecting the output

### Proportion of variance

```{r}
summary(pr_obj)
```

-   **Standard deviation** — $\sqrt{\lambda_k}$.
-   **Proportion of Variance** — $\lambda_k / \sum_i \lambda_i$, i.e.
    how much of the total variance each component explains.
-   **Cumulative Proportion** — running total of the above.

PC1 explains about 97.5% of total variance. Retaining just one component
is clearly sufficient here.

### Scree plot

```{r}
plot(pr_obj, main = "Scree plot – Painted Turtles")
```

Plots the raw eigenvalues $\lambda_k$ against component number. Look for
an elbow — the point after which eigenvalues drop off sharply. Here the
drop after PC1 makes the choice obvious.

### Scores

```{r}
head(pr_obj$x)
```

`pr_obj$x` contains the PC scores $\mathbf{Y} = \mathbf{X}\mathbf{V}$ —
each row is one observation's coordinates in the PC space.

------------------------------------------------------------------------

## 4. From first principles using the SVD

### Step 1: Prepare the data matrix

```{r}
data_turtle_mat <- data_tidy_turtle |>
  dplyr::select(-gender) |>
  dplyr::mutate(dplyr::across(dplyr::everything(), log)) |>
  dplyr::mutate(dplyr::across(dplyr::everything(),
                               function(x) x - mean(x))) |>
  as.matrix()

head(round(data_turtle_mat, 4))
```

We manually do what `prcomp` does internally:

1.  Remove the non-numeric `gender` column.
2.  Log-transform all variables.
3.  Centre each variable by subtracting its mean.

### Step 2: SVD

```{r}
svd_obj <- svd(data_turtle_mat)
```

This returns:

-   `svd_obj$d` — singular values $d_1 \geq d_2 \geq \cdots \geq d_p$.
-   `svd_obj$u` — left singular vectors ($n \times p$).
-   `svd_obj$v` — right singular vectors ($p \times p$), these are the
    **loading vectors**.

### Step 3: Eigenvalues of the covariance matrix

The eigenvalues of $\mathbf{S} = \mathbf{X}'\mathbf{X}/(n-1)$ are

$$\lambda_k = d_k^2 / (n - 1).$$

```{r}
n <- nrow(data_turtle_mat)
eigenvalues <- svd_obj$d^2 / (n - 1)
round(eigenvalues, 4)
```

### Step 4: Loading vectors

```{r}
# Columns of V are the loading vectors a_1, a_2, ..., a_p
round(svd_obj$v, 4)
```

These match `pr_obj$rotation` up to possible sign flips (the sign of an
eigenvector is arbitrary).

### Step 5: PC scores

```{r}
# Y = XV
scores <- data_turtle_mat %*% svd_obj$v
head(round(scores, 4))
```

Compare with `pr_obj$x` — identical up to sign flips.

### Step 6: Proportion of variance explained

```{r}
prop_var <- eigenvalues / sum(eigenvalues)
round(prop_var, 4)

## Alternative scree plot (hsowing prop var)
plot(prop_var, type = 'b')

cumsum(round(prop_var, 4))
```

------------------------------------------------------------------------

## 5. Biplot

```{r}
biplot(pr_obj, main = "Biplot – Painted Turtles")
```

`biplot()` takes the `prcomp` object directly and plots:

-   **Points** — observations (rows of $\mathbf{Y}$, the scores).
-   **Arrows** — variables (columns of $\mathbf{V}$, the loadings).

Two rules for interpretation:

1.  An observation near the **tip of a variable's arrow** is above
    average on that variable.
2.  Variables with arrows pointing in **similar directions** are
    correlated; arrows that are **orthogonal** indicate uncorrelated
    variables.

Here all three arrows point in roughly the same direction, confirming
the high correlations. PC1 separates small turtles (left) from large
ones (right).

------------------------------------------------------------------------

## Summary of key objects

| Object | What it is | How to get it |
|----|----|----|
| $\mathbf{V}$ | Loading matrix | `pr_obj$rotation` or `svd_obj$v` |
| $\lambda_k$ | Eigenvalues | `pr_obj$sdev^2` or `svd_obj$d^2 / (n-1)` |
| $\mathbf{Y}$ | Scores | `pr_obj$x` or `data_mat %*% svd_obj$v` |
| $\hat{\mathbf{X}}$ | Rank-$s$ approximation | `scores[,1:s] %*% t(svd_obj$v[,1:s])` |
