---
title: "CA10"
format: pdf
---

# Question 1

```{r}
if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
if (!requireNamespace("DataTidy23RodoHonsMult", quietly = TRUE)) {
# note that you'll need to reinstall the package
# to get the latest version (as of 24 April 2024)
# with the data set
remotes::install_github("MiguelRodo/DataTidy23RodoHonsMult@2024")
}
data("corr_mat_profit", package = "DataTidy23RodoHonsMult")
corr_mat_profit
```

```{r}
#Accounting Measures
X1 = corr_mat_profit[1:6, 1:6]

#Market Value 
X2 = corr_mat_profit[7:8, 7:8]

X21  = corr_mat_profit[1:6, 7:8]

X12 = corr_mat_profit[7:8, 1:6]
```

## Perform CCA

```{r}
X1_inv = solve(X1)

X1.eig = eigen(X1_inv)

X1_inv.sqrt = X1.eig$vectors %*% diag(sqrt(X1.eig$values)) %*% t(X1.eig$vectors)
```

```{r}
X2_inv = solve(X2)

X2.eig = eigen(X2_inv)

X2_inv.sqrt = X2.eig$vectors %*% diag(sqrt(X2.eig$values)) %*% t(X2.eig$vectors)
```

### Eigen decomposition

```{r}
rho = X1_inv.sqrt %*% X21 %*% X2_inv %*% X12 %*% X1_inv.sqrt

rho_eigen = eigen(rho)
rho_eigen
```

### Get Loadings ( Question 1a)

```{r}
#For X1
a1 = X1_inv.sqrt %*% rho_eigen$vectors[ ,1]
a1

#For X2 
b1 = X2_inv %*% X12 %*% a1
b1
```

### Question 1c

```{r}
#Calc covariance for set 2 (b1) and from diagonal mat of variances
# V = 1/(sqrt(t(a1) %*% X1 %*% a1))
# V = matrix(c(V, V),2,1)
# V
# 
# #Standardize to unit var 
# a1 = a1 * V
# a1 
# 
# #Calc canonical correlation
# rho_1 = sqrt(rho_eigen$values[1])
# rho_1
```

```{r}
#Calc covariance for set 2 (b1) and from diagonal mat of variances
V = 1/(sqrt(t(b1) %*% X2 %*% b1))
V = matrix(c(V, V),2,1)
V

#Standardize to unit var 
b1 = b1 * V
b1 

#Calc canonical correlation
rho_1 = sqrt(rho_eigen$values[1])
rho_1
```

### Question 1d

Interpret canonical variates using the coefficient vectors.

### Question 1e

Interpret canonical variables using correlation matrices for each set

# Question 2

```{r}
sigma = matrix(c(100, 0,0,0,0,1,0.95,0,0,0.95,1,0,0,0,0,100), 4, 4, byrow = T)
sigma 
```

### Question 2a

CCA does not choose variables to include in the ,inear combinations based on maximizing variance, it does so by maximizing correlations. (Look at off diagonal blocks to see the correlations)

### Question 2b

0.  It mean that there exists no correlation between the sets linear combinations

### Question 2c

It affects the scaling of the canonical variables, as well as the correlations between the canonical variables.

### Question 2d

(Affected by what?)

# Question 3
