---
title: "Multivariate Analysis: Lecture 4"
format: html
editor: visual
---

## Exercise 4.3

### Question 1

```{r}
library(readr)
Air_Pollution_Data <- read_csv("C:/Users/isabe/OneDrive - University of Cape Town/Documents/GitHub/Multivariate-Analysis/Multivariate_Analysis/Lecture_1/Air Pollution Data.csv")
View(Air_Pollution_Data)

#Construct a QQ-plot 
sol <- na.omit(Air_Pollution_Data$Radiation)
qqnorm(sol) ; qqline(sol, col = "red", lwd = 2)

#Number of values to sum over
n <- length(sol)

#Ordered sample values
x <- sort(sol)
length(x)

#Theoretical normal quantiles with approximation 
q <- qnorm((1:n - 0.5)/n)
length(q)

#Determine correlation coefficient 
rq_formula <- sum((x - mean(x))*(q - mean(q)))/(sqrt(sum((x - mean(x))^2))*sqrt(sum((q - mean(q))^2)))
rq_formula <- round(rq_formula, 4)

#Use built in R formula to check 
rq <- cor(x,q)

#Check sample size 
length(sol)
```
There are 42 observations. The correlation coefficient from the above test is `r rq_formula`. The critical value in Table B.1 for a sample size of 40 is 0.9726 for alpha = 0.05. Since `r rq_forumala` is < the critical value of 0.9726 we reject the assumption of normality at a 5% level. The critical value for a smaple size of 45 is 0.9749 for alpha = 0.05. 

At a 10% level our correlation coefficient is greater than both critcial values for a sample size of 40 and 45, 0.9599 and 0.9632, respectively. We would not reject the assumption of normality at a 10% level. 

It would make sense to reject the assumption of normality after looking at the QQ-plot as the data points do not lie on the normal line.

### Question 2
```{r}
#Extract covariates we are interested in 
x5 <- Air_Pollution_Data$NO2
x6 <- Air_Pollution_Data$O3
x <- cbind(x5, x6)

#S: covariance matrix 
S <- cov(x)

#n: number of observations 
n <- nrow(x)

#Calculate squared distances 
d2 <- mahalanobis(x, colMeans(x), cov = S)

#Order the squared distances
d2_ord <- sort(d2)

#Determine 50% cirtical value
crit_val <- qchisq(0.5, df = 2)

#determine how many d2 values are less than critical value 
prop <- mean(d2_ord < crit_val)
prop*100

#Graph the pairs
qcp <- qchisq((1:n - 0.5)/n, ncol(x))
plot(qcp, d2_ord, main="Chi-square plot", pch = 16, 
     ylab = '') ; title(ylab = expression(d[(j)]^2), line = 2) ; abline(a = 0, b = 1, col = 'red')


```