---
title: "practice7"
format: html
---

```{r}
library(dplyr)
library(knitr)
```

```{r}
if (!requireNamespace("remotes", quietly = TRUE)) {
install.packages("remotes")
}
# be sure to install from `2024` branch (code below does this)
remotes::install_github("MiguelRodo/DataTidy23RodoHonsMult@2024")
data("data_tidy_mali", package = "DataTidy23RodoHonsMult")
```

# Q1

```{r}
dat = data_tidy_mali
plot(dat$family ~ dat$dist_rd)
plot(dat$dist_rd ~ dat$cattle)

dat = dat %>% filter(dat$cattle < 40, dat$dist_rd < 200, dat$family < 100)

plot(dat$family ~ dat$dist_rd)
plot(dat$dist_rd ~ dat$cattle)

```

# Q2

We need to standardise as the predictors are on very different scales

```{r}
dat = scale(dat)
```

# Q3

```{r}
cor_mat = cor(dat)
eigen = eigen(cor_mat)

total_var = sum(eigen$values)
prop = eigen$values / total_var

plot(prop, type = 'b', xlab = "PCs", ylab = "Prop var", main = "Scree plot")
```

We choose 2 PCs

```{r}
PC1 = eigen$vectors[,1]
PC2 = eigen$vectors[,2]
PCs = data.frame("PC1" = PC1, "PC2" = PC2)
kable(PCs)

score1 = dat %*% PC1
score2 = dat %*% PC2
scores = data.frame("Score1" = score1, "Score2" = score2)
kable(scores)
```

# Q4

```{r}
# Note so either take eigen(cov) or svd(data)
svd_result = svd(dat)
loadings = svd_result$v
df_loading = data.frame(loadings[,1:2])
kable(df_loading)

alt_PCscore = dat %*% svd_result$v
data.frame("score1" = alt_PCscore[,1], "score2" = alt_PCscore[,2])

scores = svd_result$u %*% diag(svd_result$d)
svd_pc = data.frame("score1" = scores[,1], "score2" = scores[,2])
kable(svd_pc)
```

# 5

nah

# 6

```{r}
# correlation is with score not loading
corPC1 = cor(score1, dat)
df = data.frame(corPC1)
kable(df)
```

We can see the PC is positively correlated with each var which makes sense as we saw that its loading from each var was positive. Additionally the vars that loaded the storngest on PC1 are the vars that are more correlated with it.

# Q7

```{r}
# Just learn prcomp much easier for this part
pca_prcomp = prcomp(dat, scale. = FALSE)

biplot(pca_prcomp, choices = c(1,2))

biplot(x = scores, y = loadings)
```

if orthogonal to PC then doesnt load on it or barely, if parallel to PC then loads a lot on it. zthen always spea baout the points too if they are clustered together which PC has more variability shown with the points etc etc.
